Revealing transient strain in geodetic data with Gaussian process regression

### 2017-10-02 RevisedSubmission/hines_hetland_2017_gji.tex
did more

Here we propose using Gaussian process regression (GPR) \citep{Rasmussen2006} to estimate transient strain from GNSS data. GPR is a Bayesian, non-parametric method for inferring a continuous signal from scattered data. Since GNSS stations are irregularly spaced and observation times may differ between stations, GPR is an ideal tool for synthesizing discrete GNSS data into a spatially and temporally continuous representation of surface deformation. GPR is closely related to kriging \citep{Cressie1992} and least squares collocation \citep{Moritz1978}. The latter has been used by \citet{Kato1998} and \citet{El-Fiky1999} to estimate secular strain rates from GNSS data. GPR is Bayesian in that we describe our prior understanding of the geophysical signal with a Gaussian process. A Gaussian process is a normally distributed stochastic process that is fully defined in terms of a mean function and a positive definite covariance function. For example, Brownian motion, $B(t)$, is a well known Gaussian process in $\mathbb{R}^1$ which has zero mean and covariance function $\mathrm{cov}(B(t),B(t')) = \min(t,t')$, where $t,t' \ge 0$. If no prior information is available for the geophysical signal, then maximum likelihood methods can be used to objectively choose a prior Gaussian process that is most consistent with the observations.  We incorporate GNSS observations with the prior to form a posterior estimate of transient strain.  The posterior transient strain is also a Gaussian process, and we can use its distribution to confidently discern geophysical signal from noise. We use GPR to infer strain resulting from SSEs in the Pacific Northwest, demonstrating that GPR is an effective tool for detecting transient geophysical processes.

\section{Estimating Transient Strain Rates}\label{sec:Method}
We seek a spatially and temporally dependent estimate of transient strain rates. We consider transient strain rates to be any deviation from secular strain rates, and our attention is limited to horizontal strain rates. We denote transient strain rates as $\dot\varepsilon(p)$, where $p$ represents the ordered pair $(\vec{x},t)$, $\vec{x}$ are spatial coordinates, and $t$ is time. For simplicity, we assume that the study region is sufficiently small that $\vec{x}$ can be considered a point in a 2-D map projection that is aligned with the cardinal directions, $\vec{x} = (x_\mathrm{e},x_\mathrm{n})$. We determine $\dot\varepsilon$ by spatially and temporally differentiating estimates of transient displacements, $\vec{u}(p)$. Transient displacements have an easting and northing component, $\vec{u} = (u_\mathrm{e}, u_\mathrm{n})$, and we make a prior assumption that each component of $\vec{u}$ is a Gaussian process with zero mean and a generic covariance function $\mathrm{Cov}\left(u_i(p),u_i(p')\right) = C_{u_i}(p,p')$. Using a more concise notation, we describe each component of $\vec{u}$ as $u_i \sim \mathcal{N}\left(0,C_{u_i}\right)$. The function $C_{u_i}$ must be positive definite, which means that the matrix $[C_{u_i}(p,p')]_{(p,p') \in \mitbf{P} \times \mitbf{P}}$ is positive definite for any set of points $\mitbf{P}$ \citep[sec. 2.5]{Cressie1992}. We assume that $C_{u_i}$ can be separated into spatial and temporal functions as 
\begin{equation}\label{eq:TransientCovariance}
C_{u_i}\left((\vec{x},t),(\vec{x}',t')\right) = X_i(\vec{x},\vec{x}')T_i(t,t').
\end{equation}  
As long as the functions $X_i$ and $T_i$ are positive definite, $C_{u_i}$ is guaranteed to also be positive definite and thus a valid covariance function \citep[sec. 4.2.4]{Rasmussen2006}. We also require that the derivatives $\partial^2 X_i(\vec{x},\vec{x}')/ \partial x_j \partial x_j'$ and $\partial^2 T_i(t,t') / \partial t \partial t'$ exist everywhere, which ensures that $u_i$ is spatially and temporally differentiable (See \citet[sec. 2.2]{Adler1981} for a formal definition of differentiation for Gaussian processes). The appropriate choice for $X_i$ and $T_i$ may vary depending on the geophysical signal we are trying to describe.  To keep this section sufficiently general, we hold off on specifying $X_i$ and $T_i$ until Section \ref{sec:SignalModel}, where we estimate transient strain from slow slip events in the Pacific Northwest.

@@

We combine the prior on transient displacements with GNSS displacement data to form a posterior estimate of transient displacements. We ignore any covariances between the easting and northing components, which reduces the computational complexity of our analysis because we can determine each component of the posterior independently. However, we are inherently assuming that the principle components of transient displacements and the noise processes are aligned with the cardinal directions, which is admittedly an arbitrary assumption. We then form each component of the posterior, denoted as $\hat{u}_i$, by conditioning $u_i$ with the fact that $\mitbf{d}^*_i$ has been realized from the random vector $\mitbf{d}_i$. That is to say $\hat{u}_i = u_i |(\mitbf{d}_i = \mitbf{d}^*_i)$. If $\kappa$ is kept finite then, from \citet[sec. 2.2]{Rasmussen2006}, we find that $\hat{u}_i$ is distributed as $\mathcal{N}(\mu_{\hat{u}_i},C_{\hat{u}_i})$, where
\begin{equation}\label{eq:PosteriorMean}
\mu_{\hat{u}_i}(p) = C_{u_i}(p,\mitbf{P})\left(\mitbf{\Sigma} + \kappa^2\mitbf{G}\mitbf{G}^T\right)^{-1}\mitbf{d}^*_i
\begin{equation}\label{eq:PosteriorCov}
C_{\hat{u}_i}(p,p') = C_{u_i}(p,p') - C_{u_i}(p,\mitbf{P})\left(\mitbf{\Sigma} + \kappa^2\mitbf{G}\mitbf{G}^T\right)^{-1}C_{u_i}(\mitbf{P},p').

@@

\mu_{||\dot\varepsilon||_\mathrm{F}} &\approx ||\dot\varepsilon||_F \big|_{\dot\varepsilon = \mu_{\dot\varepsilon}} \\

@@

\subsection{Outlier detection}\label{sec:Outlier}
In deriving our formulation for transient strain rates, we have assumed that noise in the data vector is normally distributed. This is not an appropriate assumption for GNSS data, which often have more outliers than would be predicted for normally distributed noise. Methods for analyzing GNSS data should either be robust against outliers or should involve a preprocessing step in which outliers are detected and removed. Examples of the former include the MIDAS method for estimating secular velocities \citep{Blewitt2016} and the GPS Imaging method for detecting spatially coherent features \citep{Hammond2016}.  We adopt the latter approach and remove outliers in the GNSS data prior to estimating $\dot\varepsilon$. We identify outliers based on the residuals between the observations and a best fitting model. Observations with residuals that exceed some threshold are removed. This strategy is commonly used to detect outliers in GNSS data, where the residuals are typically computed for each station by estimating and removing a temporally dependent model consisting of a linear trend and seasonal terms \citep[e.g.,][]{Johansson2002,Dong2006,Bos2013}. For the outlier detection method used in this study, the residuals are determined by removing a best fitting spatially and temporally dependent model from the GNSS data. That model that we want to fit to the $n$ GNSS observations, $\mitbf{d}^*$, consists of the linear trend and sinusoids from eq. (\ref{eq:DataVec}), $\mitbf{G}\mitbf{m}$, and a Gaussian process, $v(\mitbf{P})$, that is used to describe any spatially and temporally correlated features in the GNSS data. We now consider $\mitbf{d}^*$ to be a realization of the random vector
\begin{equation}
\mitbf{d} = \mitbf{G}\mitbf{m} + v(\mitbf{P}) + \mitbf{w},
\end{equation}   
where $\mitbf{w}$ is a vector of uncorrelated Gaussian noise distributed as $\mitbf{w} = \mathcal{N}\left(0,\mathrm{diag}(\mitbf{\sigma}^2)\right)$, and $\mitbf{\sigma} = [\sigma_1, \sigma_2, ..., \sigma_n]$ are the data uncertainties. The residual vector, $\mitbf{r} = [r_1,r_2,...,r_n]$, is the difference between $\mitbf{d}^*$, and the mean of the $\mitbf{G}\mitbf{m} + v(\mitbf{P})$ conditioned on $\mitbf{d} = \mitbf{d}^*$.

>>>>

Here we propose using Gaussian process regression (GPR) \citep{Rasmussen2006} to estimate transient strain from GNSS data. GPR is a Bayesian, non-parametric method for inferring a continuous signal from scattered data. Since GNSS stations are irregularly spaced and observation times may differ between stations, GPR is an ideal tool for synthesizing discrete GNSS data into a spatially and temporally continuous representation of surface deformation. GPR is closely related to kriging \citep{Cressie1992} and least squares collocation \citep{Moritz1978}. The latter has been used by \citet{Kato1998} and \citet{El-Fiky1999} to estimate secular strain rates from GNSS data. GPR is Bayesian in that we describe our prior understanding of the geophysical signal with a Gaussian process. A Gaussian process is a normally distributed stochastic process that is fully defined in terms of a mean function and a positive definite covariance function. For example, Brownian motion, $B(t)$, is a well known Gaussian process in $\mathbb{R}^1$ which has zero mean and covariance function $\mathrm{Cov}\left[B(t),B(t')\right] = \min(t,t')$, where $t,t' \ge 0$. If no prior information is available for the geophysical signal, then maximum likelihood methods can be used to objectively choose a prior Gaussian process that is most consistent with the observations.  We incorporate GNSS observations with the prior to form a posterior estimate of transient strain.  The posterior transient strain is also a Gaussian process, and we can use its distribution to confidently discern geophysical signal from noise. We use GPR to infer strain resulting from SSEs in the Pacific Northwest, demonstrating that GPR is an effective tool for detecting transient geophysical processes.

\section{Estimating Transient Strain Rates}\label{sec:Method}
We seek a spatially and temporally dependent estimate of transient strain rates. We consider transient strain rates to be any deviation from secular strain rates, and our attention is limited to horizontal strain rates. We denote transient strain rates as $\dot\varepsilon(p)$, where $p$ represents the ordered pair $(\vec{x},t)$, $\vec{x}$ are spatial coordinates, and $t$ is time. For simplicity, we assume that the study region is sufficiently small that $\vec{x}$ can be considered a point in a 2-D map projection that is aligned with the cardinal directions, $\vec{x} = (x_\mathrm{e},x_\mathrm{n})$. We determine $\dot\varepsilon$ by spatially and temporally differentiating estimates of transient displacements, $\vec{u}(p)$. Transient displacements have an easting and northing component, $\vec{u} = (u_\mathrm{e}, u_\mathrm{n})$, and we make a prior assumption that each component of $\vec{u}$ is a Gaussian process with zero mean and a generic covariance function $\mathrm{Cov}\left[u_i(p),u_i(p')\right] = C_{u_i}(p,p')$. Using a more concise notation, we describe each component of $\vec{u}$ as $u_i \sim \mathcal{N}\left(0,C_{u_i}\right)$. The function $C_{u_i}$ must be positive definite, which means that the matrix $[C_{u_i}(p,p')]_{(p,p') \in \mitbf{P} \times \mitbf{P}}$ is positive definite for any set of points $\mitbf{P}$ \citep[sec. 2.5]{Cressie1992}. We assume that $C_{u_i}$ can be separated into spatial and temporal functions as 
\begin{equation}\label{eq:TransientCovariance}
C_{u_i}\left((\vec{x},t),(\vec{x}',t')\right) = X_i(\vec{x},\vec{x}')T_i(t,t').
\end{equation}  
As long as the functions $X_i$ and $T_i$ are positive definite, $C_{u_i}$ is guaranteed to also be positive definite and thus a valid covariance function \citep[sec. 4.2.4]{Rasmussen2006}. We also require that the derivatives $\partial^2 X_i(\vec{x},\vec{x}')/ \partial x_j \partial x_j'$ and $\partial^2 T_i(t,t') / \partial t \partial t'$ exist everywhere, which ensures that $u_i$ is spatially and temporally differentiable (See \citet[sec. 2.2]{Adler1981} for a formal definition of differentiation for Gaussian processes). The appropriate choice for $X_i$ and $T_i$ may vary depending on the geophysical signal we are trying to describe.  To keep this section sufficiently general, we hold off on specifying $X_i$ and $T_i$ until Section \ref{sec:SignalModel}, where we estimate transient strain from slow slip events in the Pacific Northwest.

@@

We combine the prior on transient displacements with GNSS displacement data to form a posterior estimate of transient displacements. We ignore any covariances between the easting and northing components, which reduces the computational complexity of our analysis because we can determine each component of the posterior independently. However, we are inherently assuming that the principle components of transient displacements and the noise processes are aligned with the cardinal directions, which is admittedly an arbitrary assumption. We then form each component of the posterior, denoted as $\hat{u}_i$, by conditioning $u_i$ with the fact that $\mitbf{d}^*_i$ has been realized from the random vector $\mitbf{d}_i$. That is to say $\hat{u}_i = u_i |(\mitbf{d}_i = \mitbf{d}^*_i)$. A general solution for $\hat{u}_i$ is derived in \citet[sec. 8.9]{VonMises1964}, where we find that $\hat{u}_i$ is distributed as $\mathcal{N}(\mu_{\hat{u}_i},C_{\hat{u}_i})$ with mean function
\begin{align}\label{eq:PosteriorMean}
\mu_{\hat{u}_i}(p) &= \mathrm{E}\left[u_i(p)\right] + 
                      \mathrm{Cov}\left[u_i(p),\mitbf{d}_i\right] 
                      \mathrm{Cov}\left[\mitbf{d}_i\right]^{-1}
                      \left(\mitbf{d}^*_i - \mathrm{E}\left[\mitbf{d}_i\right]\right)\nonumber \\
                   &= C_{u_i}(p,\mitbf{P})\left(\mitbf{\Sigma} + \kappa^2\mitbf{G}\mitbf{G}^T\right)^{-1}\mitbf{d}^*_i
and covariance function
\begin{align}\label{eq:PosteriorCov}
C_{\hat{u}_i}(p,p') &= \mathrm{Cov}\left[u_i(p),u_i(p')\right] - 
                       \mathrm{Cov}\left[u_i(p),\mitbf{d}_i\right] 
                       \mathrm{Cov}\left[\mitbf{d}_i\right]^{-1}
                       \mathrm{Cov}\left[\mitbf{d}_i,u_i(p')\right]\nonumber \\
                    &= C_{u_i}(p,p') - C_{u_i}(p,\mitbf{P})\left(\mitbf{\Sigma} + \kappa^2\mitbf{G}\mitbf{G}^T\right)^{-1}C_{u_i}(\mitbf{P},p').

@@

\mu_{||\dot\varepsilon||_\mathrm{F}} &\approx ||\dot\varepsilon||_F \big|_{\dot\varepsilon = \mu_{\dot\varepsilon}}\nonumber \\

@@

\subsection{Outlier detection}\label{sec:Outlier}
In deriving our formulation for transient strain rates, we have assumed that noise in the data vector is normally distributed. This is not an appropriate assumption for GNSS data, which often have more outliers than would be predicted for normally distributed noise. Methods for analyzing GNSS data should either be robust against outliers or should involve a preprocessing step in which outliers are detected and removed. Examples of the former include the MIDAS method for estimating secular velocities \citep{Blewitt2016} and the GPS Imaging method for detecting spatially coherent features \citep{Hammond2016}.  In this study, we identify and remove outliers as a preprocessing step before estimating $\dot\varepsilon$. Outliers are identified based on the residuals for a model that best fits the observed data. Observations with residuals that exceed some threshold are removed. This strategy is commonly used to detect outliers in GNSS data, where the residuals are typically computed by estimating and removing a linear trend and seasonal terms for each time series \citep[e.g.,][]{Johansson2002,Dong2006,Bos2013}. The outlier detection method used here is different from this more standard approach because the model we fit to the data additionally consists of a spatially and temporally correlated Gaussian process. With this more complex model, we are able to detect outliers that are spatially or temporally anomalous.

For the outlier detection method used in this study, the residuals are determined by removing a best fitting spatially and temporally dependent model from the GNSS data. That model that we want to fit to the $n$ GNSS observations, $\mitbf{d}^*$, consists of the linear trend and sinusoids from eq. (\ref{eq:DataVec}), $\mitbf{G}\mitbf{m}$, and a Gaussian process, $v(\mitbf{P})$, that is used to describe any spatially and temporally correlated features in the GNSS data. We now consider $\mitbf{d}^*$ to be a realization of the random vector
\begin{equation}
\mitbf{d} = \mitbf{G}\mitbf{m} + v(\mitbf{P}) + \mitbf{w},
\end{equation}   
where $\mitbf{w}$ is a vector of uncorrelated Gaussian noise distributed as $\mitbf{w} = \mathcal{N}\left(0,\mathrm{diag}(\mitbf{\sigma}^2)\right)$, and $\mitbf{\sigma} = [\sigma_1, \sigma_2, ..., \sigma_n]$ are the data uncertainties. The residual vector, $\mitbf{r} = [r_1,r_2,...,r_n]$, is the difference between $\mitbf{d}^*$, and the mean of the $\mitbf{G}\mitbf{m} + v(\mitbf{P})$ conditioned on $\mitbf{d} = \mitbf{d}^*$.
